\section{Discussion}

Our measurements reveal a clear hierarchy in how different Android configurations handle user data and privacy. The most fundamental finding is that adjusting privacy settings on stock Android, while beneficial, cannot provide the same level of protection as switching to an alternative distribution. On systems running proprietary Google services, privacy-adjusted scenarios showed only modest reduction in data transmission compared to the default setup. This is due to the underlying architecture which tightly integrates Google components into core of OS and maintains persistent background activity. As a result, substantial data flows to Google infrastructure continue regardless of user-facing privacy settings.

Alternative distributions shift the privacy posture by making Google communication user-initiated rather than ambient. All privacy-focused systems generated orders of magnitude less network traffic with Google. Some of them utilize their own or third-party privacy-focused infrastructure to replace Google services. LineageOS with GMS was an exception, while more minimal, by having official Google services preinstalled, it does not differ much from stock Android in terms of privacy. Among other configurations enabling basic Google functionality, microG represents the most privacy-preserving approach, though this comes with potential security trade-offs (signature spoofing, privileged install). Our measurements show that microG transmits the least data to Google infrastructure while maintaining basic service compatibility. It achieves this through identifier randomization rather than exposing authentic hardware identifiers, similar to GrapheneOS. Our observations illustrate the fundamental difference: microG optimizes for privacy, whereas GrapheneOS prioritizes security, but still providing strong privacy, substantially improved over stock systems.

Alternative distributions also improve user control through minimal base systems and enhanced permission granularity. All tested alternatives ship without forced user agreements during initial setup, unlike stock Android, which requires accepting mandatory terms including data collection. A notable privacy-enhancing feature present in all tested systems, which is absent in stock Android, is per-application network permission control, allowing users to restrict Internet access for specific applications. Aurora Store demonstrates that downloading applications from official repositories remains viable without full Google infrastructure interaction, retrieving identical packages from the same source while generating fewer requests. Updates in alternative systems are typically smaller, more transparent, and user-controlled rather than automatic and opaque.

Several limitations qualify these findings. Our measurements capture specific, time-bounded scenarios rather than universal behavior. Traffic patterns likely vary with extended use and evolving service configurations. The TLS interception approach, while revealing substantial information, remains incomplete due to certificate pinning, anti-tamper mechanisms, and ongoing protocol evolution. GrapheneOS's resistance to our instrumentation, while security-advantageous, means we cannot closely verify what data sandboxed services transmit. These constraints indicate that our findings reflect general trends and architectural differences rather than definitive or absolute measurements.

Based on the network traffic we analyzed, Google may potentially distinguish between stock Android and alternative distributions through device characteristics transmitted during service interactions, raising the possibility of selective enforcement or access restrictions. Ultimately, the choice between privacy-focused alternative operating systems depends on individual threat models and functional requirements, but all of them represent meaningful improvements in user control and privacy over stock Android.

\section{Conclusion}

In this work, we compared network behavior across multiple Android distributions, measuring data transmission during setup, idle periods, and basic application usage. We evaluated Stock Android, LineageOS variants, iodéOS, and GrapheneOS to quantify how different approaches to Google services affect user privacy. Our findings show that alternative distributions reduce unsolicited data flows by orders of magnitude compared to stock systems. These systems enhance privacy by reducing data exposure, limiting or randomizing identifiers, and using independent infrastructure. Network traffic and Google Takeout analysis revealed that stock Android transmits persistent hardware identifiers and various telemetry data, whereas alternative systems either do not transmit such data or significantly limit its scope.

Future work could extend measurements to longer usage periods and repeated trials to distinguish systematic differences from transient variation. More realistic tasks incorporating diverse applications would better approximate everyday use. As Android, Google services and their alternatives continue to evolve, repeating this study would help determinate whether future privacy improvements meaningfully alter the patterns we observed.

